### 2020-07-20 paper.tex
tweaks

\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
@gen function q(trace)
  current_k = trace[:k]
  is_birth ~ bernoulli(current_k == 1 ? 1.0 : 0.5)
  if is_birth
    new_mu ~ normal(0, 10)
    new_var ~ inv_gamma(1, 10)
  else
    deletion_idx ~ uniform_discrete(1, current_k)
  end
end

@transform h (tr1, tr2) to (tr3, tr4) begin
   is_birth = @read(tr2, :is_birth, :discrete)
   @write(tr4[:is_birth], !is_birth, :discrete)
   k = @read(tr1[:k], :discrete)
   weights = @read(tr1[:weights], :continuous)
   if is_birth
     @write(tr3[:k], k+1, :discrete)
     new_mu = @read(tr2[:new_mu], :continuous)
     new_var = @read(tr2[:new_var], :continuous)
     @write(tr3[(:mu, k+1)], new_mu, :continuous)
     @write(tr3[(:var, k+1)], new_var, :continuous)
     new_weights = add_weight(weights)
     @write(tr4[:deletion_idx], k+1, :discrete)
   else
     idx = @read(tr2[:deletion_idx], :discrete)
     @copy(tr1[(:mu, idx)], tr4[:new_mu])
     @copy(tr1[(:var, idx)], tr4[:new_var])
     for i in (idx+1):k
       @copy(tr1[(:mu, i)], tr3[(:mu, i-1)])
       @copy(tr1[(:var, i)], tr3[(:var, i-1)])
     end
     @write(tr3[:k], k-1, :discrete)
     new_weights = delete_weight(weights, idx)
   end
   @write(tr3[:weights], new_weights, :continuous)
end
\end{lstlisting}}
\end{tabular}
\end{center}

@@

\paragraph{Hierarchical address spaces.}
This example uses probabilistic programs that invoke other probabilistic programs, sometimes recursively.
For example, the model probabilistic program $\mathtt{p}$ invokes the probabilistic program $\mathtt{cov\_function\_prior}$, which is itself recursive.
Similarly, the auxiliary probabilistic program $\mathtt{q}$ invokes $\mathtt{walk\_tree}$ (which is recursive) as well as $\mathtt{cov\_function\_prior}$.
Consider the syntax used to recursively invoke $\mathtt{walk\_tree}$ within $\mathtt{walk\_tree}$:
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
({:left} ~ walk_tree(node.left, path))
\end{lstlisting}}
\end{tabular}
\end{center}
This expression resembles a random choice expression.
However, instead of associating the return value of the function $\mathtt{walk\_tree}$ with the address $\mathtt{:}\mathtt{left}$, the address $\mathtt{:}\mathtt{left}$ is associated with the entire trace of random choices made within $\mathtt{walk\_tree}$.
That is, $\mathtt{:}\mathtt{left}$ is the \emph{namespace} for the addresses of all random choice made within the invocation.
Further invocations by the callee themselves result in nested namespaces.
This process results in a \emph{hierarchical address space} for random choices, as shown in Figure~\ref{fig:structure-learning}b.
This does not modify the mathematical formalism---each random choice made during the execution of a probabilistic program still has a unique address, but the address has multiple components that localize it within the hierarchy.
For example, the choice $\mathtt{done}$ a recursive call to $\mathtt{walk\_tree}$ might have address:
\[
k = (\mathtt{:}\mathtt{left} \mbox{ \texttt{=>} } \mathtt{:}\mathtt{right} \mbox{ \texttt{=>} } \mathtt{:}\mathtt{done})
\]
(In Gen, `\texttt{=>}' constructs hierarchical addresses.)

\paragraph{A complex state-dependent distribution.}
At each iteration of the MCMC algorithm, the auxiliary probabilistic program $\mathtt{q}$ (Figure~\ref{fig:structure-learning}d) first picks a random node in the parse tree of the covariance function, by doing a stochastic walk of the existing parse tree that terminates at the chosen node.
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
prev_cov_function = trace[:cov_function]
path ~ walk_tree(prev_cov_function, ..)
\end{lstlisting}}
\end{tabular}
\end{center}
The code that walks the tree uses the following recursion, which results in a probability distribution that assigns exponentially lower probability to nodes that are deeper in the tree.
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
if ({:done} ~ bernoulli(0.5))
  return path
elseif ({:recurse_left} ~ bernoulli(0.5))
  path = (path..., :left_node)
  return ({:left} ~ walk_tree(node.left, path))
else
  path = (path..., :right_node)
  return ({:right} ~ walk_tree(node.right, path))
end
\end{lstlisting}}
\end{tabular}
\end{center}
The resulting distributions on selected nodes for two example trees are shown below:
\begin{center}
\includegraphics[width=0.8\linewidth]{figures/biased-tree-distribution.pdf} % TODO improve quality
\end{center}
The first part of the involution (Figure~\ref{fig:structure-learning}e) copies the random choices made during this walk from the input auxiliary trace to the output auxiliary trace.
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
@copy(aux_in[:path], aux_out[:path])
\end{lstlisting}}
\end{tabular}
\end{center}
Note that here, $\mathtt{@copy}$ is being used to copy the entire \emph{set} of random choices from the namespace $\mathtt{:}\mathtt{path}$ in $\mathtt{aux\_in}$ to the namespace $\mathtt{:}\mathtt{path}$ in $\mathtt{aux\_out}$.

Because the mixture distribution is specified using a probabilistic program, it is straightforward to modify the program $\mathtt{p}$ to define a different mixture distribution.
The code below specifies a mixture distribution that is uniform over all nodes in the tree.
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
n1 = size(node.left); n2 = size(node.right)
if ({:done} ~ bernoulli(1 / (1 + n1 + n2)))
  return path
elseif ({:recurse_left} ~ bernoulli(n1 / (n1+n2))
  path = (path..., :left_node)
  return ({:left} ~ walk_tree(node.left, path))
else
  path = (path..., :right_node)
  return ({:right} ~ walk_tree(node.right, path))
end
\end{lstlisting}}
\end{tabular}
\end{center}
The resulting distributions, for two input trees, are:
\begin{center}
\includegraphics[width=0.8\linewidth]{figures/unbiased-tree-distribution.pdf}
\end{center}
Note that the probability of choosing a given subtree to propose to is itself changed when the subtree changes.
Therefore, the mixture probabilities do not in general cancel in the the acceptance probability calculation, and must be accounted for.
For the original mixture distribution, the ratio of mixture probabilities is either $1$, $0.5$, or $2$ depending on whether the previous and new subtrees are leaf or internal nodes.
For this alternative mixture distribution, the ratio of mixture probabilities is the ratio of sizes of the two trees (e.g. $9/5$ or $5/9$ for the trees above).
In both cases, our automated involutive MCMC algorithm automatically computes the acceptance probability.

>>>>

\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
@gen function q(trace)
  current_k = trace[:k]
  is_birth ~ bernoulli(current_k == 1 ? 1.0 : 0.5)
  if is_birth
    new_mu ~ normal(0, 10)
    new_var ~ inv_gamma(1, 10)
  else
    deletion_idx ~ uniform_discrete(1, current_k)
  end
end
\end{lstlisting}}
\end{tabular}
\end{center}

\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
@transform h (tr1, tr2) to (tr3, tr4) begin
   is_birth = @read(tr2, :is_birth, :discrete)
   @write(tr4[:is_birth], !is_birth, :discrete)
   k = @read(tr1[:k], :discrete)
   weights = @read(tr1[:weights], :continuous)
   if is_birth
     @write(tr3[:k], k+1, :discrete)
     new_mu = @read(tr2[:new_mu], :continuous)
     new_var = @read(tr2[:new_var], :continuous)
     @write(tr3[(:mu, k+1)], new_mu, :continuous)
     @write(tr3[(:var, k+1)], new_var, :continuous)
     new_weights = add_weight(weights)
     @write(tr4[:deletion_idx], k+1, :discrete)
   else
     idx = @read(tr2[:deletion_idx], :discrete)
     @copy(tr1[(:mu, idx)], tr4[:new_mu])
     @copy(tr1[(:var, idx)], tr4[:new_var])
     for i in (idx+1):k
       @copy(tr1[(:mu, i)], tr3[(:mu, i-1)])
       @copy(tr1[(:var, i)], tr3[(:var, i-1)])
     end
     @write(tr3[:k], k-1, :discrete)
     new_weights = delete_weight(weights, idx)
   end
   @write(tr3[:weights], new_weights, :continuous)
end
\end{lstlisting}}
\end{tabular}
\end{center}

@@

\paragraph{Hierarchical address spaces.}
This example uses probabilistic programs that invoke other probabilistic programs, sometimes recursively.
For example, the model probabilistic program $\mathtt{p}$ invokes the probabilistic program $\mathtt{cov\_function\_prior}$, which is itself recursive.
Similarly, the auxiliary probabilistic program $\mathtt{q}$ invokes $\mathtt{walk\_tree}$ (which is recursive) as well as $\mathtt{cov\_function\_prior}$.
Consider the syntax used to recursively invoke $\mathtt{walk\_tree}$ within $\mathtt{walk\_tree}$:
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
({:left} ~ walk_tree(node.left, path))
\end{lstlisting}}
\end{tabular}
\end{center}
This expression resembles a random choice expression.
However, instead of associating the return value of the function $\mathtt{walk\_tree}$ with the address $\mathtt{:}\mathtt{left}$, the address $\mathtt{:}\mathtt{left}$ is associated with the entire trace of random choices made within $\mathtt{walk\_tree}$.
That is, $\mathtt{:}\mathtt{left}$ is the \emph{namespace} for the addresses of all random choice made within the invocation.
Further invocations by the callee themselves result in nested namespaces.
This process results in a \emph{hierarchical address space} for random choices, as shown in Figure~\ref{fig:structure-learning}b.
This does not modify the mathematical formalism---each random choice made during the execution of a probabilistic program still has a unique address, but the address has multiple components that localize it within the hierarchy.
For example, the choice $\mathtt{done}$ a recursive call to $\mathtt{walk\_tree}$ might have address:
\[
k = (\mathtt{:}\mathtt{left} \mbox{ \texttt{=>} } \mathtt{:}\mathtt{right} \mbox{ \texttt{=>} } \mathtt{:}\mathtt{done})
\]
(`\texttt{=>}' is the Gen syntax for constructing hierarchical addresses).

\paragraph{A complex state-dependent distribution.}
At each iteration of the MCMC algorithm, the auxiliary probabilistic program $\mathtt{q}$ (Figure~\ref{fig:structure-learning}d) first picks a random node in the parse tree of the covariance function, by doing a stochastic walk of the existing parse tree that terminates at the chosen node.
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
prev_cov_function = trace[:cov_function]
path ~ walk_tree(prev_cov_function, ..)
\end{lstlisting}}
\end{tabular}
\end{center}
The code that walks the tree uses the following recursion, which results in a probability distribution that assigns exponentially lower probability to nodes that are deeper in the tree.
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
if ({:done} ~ bernoulli(0.5))
  return path
elseif ({:recurse_left} ~ bernoulli(0.5))
  path = (path..., :left_node)
  return ({:left} ~ walk_tree(node.left, path))
else
  path = (path..., :right_node)
  return ({:right} ~ walk_tree(node.right, path))
end
\end{lstlisting}}
\end{tabular}
\end{center}
The resulting distributions on selected nodes for two possible input trees are shown below:
\begin{center}
\includegraphics[width=0.8\linewidth]{figures/biased-tree-distribution.pdf} % TODO improve quality
\end{center}
The first part of the involution (Figure~\ref{fig:structure-learning}e) copies the random choices made during this walk from the input auxiliary trace to the output auxiliary trace.
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
@copy(aux_in[:path], aux_out[:path])
\end{lstlisting}}
\end{tabular}
\end{center}
Note that here, $\mathtt{@copy}$ is being used to copy the entire \emph{set} of random choices from the namespace $\mathtt{:}\mathtt{path}$ in $\mathtt{aux\_in}$ to the namespace $\mathtt{:}\mathtt{path}$ in $\mathtt{aux\_out}$.

Because the mixture distribution is specified using a probabilistic program, it is straightforward to modify the program $\mathtt{p}$ to define a different mixture distribution.
The code below specifies a mixture distribution that is uniform over all nodes in the tree.
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
n1 = size(node.left); n2 = size(node.right)
if ({:done} ~ bernoulli(1 / (1 + n1 + n2)))
  return path
elseif ({:recurse_left} ~ bernoulli(n1 / (n1+n2))
  path = (path..., :left_node)
  return ({:left} ~ walk_tree(node.left, path))
else
  path = (path..., :right_node)
  return ({:right} ~ walk_tree(node.right, path))
end
\end{lstlisting}}
\end{tabular}
\end{center}
The resulting distributions, for two possible input trees, are:
\begin{center}
\includegraphics[width=0.8\linewidth]{figures/unbiased-tree-distribution.pdf}
\end{center}
Note that the probability of choosing a given subtree to propose to is itself changed when the subtree changes.
Therefore, the mixture probabilities do not in general cancel in the the acceptance probability calculation, and must be accounted for.
For the original mixture distribution, the ratio of mixture probabilities is either $1$, $0.5$, or $2$ depending on whether the previous and new subtrees are leaf or internal nodes.
For this alternative mixture distribution, the ratio of mixture probabilities is the ratio of sizes of the two trees (e.g. $9/5$ or $5/9$ for the trees above).
In both cases, our automated involutive MCMC algorithm automatically computes the acceptance probability.
